---
title: "Mathematical Theory of Finance Yield Curve"
execute: 
  echo: false
author: "Shmell Creon"
format: pdf 
editor: visual
---

```{r}
library(matlib)
library(knitr)
price_data <- read.csv("MToF Bond Prices.csv")
bond_data <- read.csv("MToF Bond Datas.csv")
bond_dates <- read.csv("MToF Bond Dates.csv")
coupons <- bond_data[[2]]
years_until_maturity <- bond_data[[3]]/365
face_value <- 100
```

# Problem 1

1.  **Why don't we just print more money?:** Bonds give the government the power to fund themselves but also act as a stronger tool for monetary policy, allowing for spending without undermining the legitimacy of the state currency by introducing excess inflation.

2.  **Yield Curve Flattening:** If long-term economic growth is uncertain but short-term is stable (possibly due to unstable political situations like new elections or instability abroad), people will bid on bonds with longer time horizons. This raises their price and brings their yield down while the short-term yield remains stable, causing the flattening.

3.  **Quantitative Easing:** Quantitative easing is a monetary policy wherein the central banks increase the money supply to stimulate market growth by buying back government bonds (and other securities). For example, the government, at its peak during Covid,  was purchasing "at least \$80 billion a month in Treasuries and \$40 billion in residential and commercial mortgage-backed securities" (Milstein & Wessel, n.d.).

# Problem 2

An ideal selection of bonds should have maturity spaced roughly at 6-month intervals so that any interpolation required between points has an equal distance and doesn't magnify any errors more than others. An even 6-month distribution across 10 stocks gives us the 0–5-year yield curve we are investigating.

We want the coupon payments to be near the date of maturity of the previous stocks since that’s the rate we will use to discount them. If two stocks have similar maturities, some other deciding factors are a more moderate coupon to avoid outliers, and a more recent issue date for a more accurate price. Using these as our north star, the following ten bonds have been chosen (Markets Insider, n.d.-a, Markets Insider, n.d.-b) :

-   `CA135087E679`: CAN 1.50 Jun 26
-   `CA135087S547`: CAN 3.00 Feb 27
-   `CA135087F825`: CAN 1.00 Jun 27
-   `CA135087T958`: CAN 2.25 Feb 28
-   `CA135087H235`: CAN 2.00 Jun 28
-   `CA135087Q491`: CAN 3.25 Sep 28
-   `CA135087J397`: CAN 2.25 Jun 29
-   `CA135087N670`: CAN 2.25 Dec 29
-   `CA135087T388`: CAN 2.75 Sep 30
-   `CA135087L443`: CAN 0.50 Dec 30

# Problem 3

The eigenvectors of the covariance matrix, sometimes called principal components, give us a direction of highest variance. Mathematically, when the data is projected onto this vector, it has the highest variance. Intuitively, one may think of the eigenvector as describing the mix of random variables that has most of the variation. In this context our eigenvector represents the weights of a portfolio that explains most of the variance in the market

The eigenvalues of these vectors are a measure of the variance captured by its respective eigenvector; a larger eigenvalue means that it has more variance. Ordering these eigenvectors by their respective eigenvalue’s magnitude, we can glean across which axes we have relative noise (small eigenvalues) vs. important data (large eigenvalues).

# Problem 4

## a) YTM Plot

```{r}
calculate_ytm <- function(price, face_value, coupon_rate, time_to_maturity) {
  
  coupon_payment <- (coupon_rate * face_value) / 2
  n_periods <- time_to_maturity * 2
  
  # Create YTM polynomial ( (price - discounted coupons)YTM^n )
  bond_equation <- function(ytm) {
    
    t <- 1:n_periods
    pv_coupons <- sum(coupon_payment / (1 + ytm/2)^t)
    pv_face <- face_value / (1 + ytm/2)^n_periods
    
    return((pv_coupons + pv_face) - price)
  }
  
  result <- uniroot(bond_equation, interval = c(-0.1, 0.5))$root
  
  return(result)
}
```

```{r, include = FALSE}
par(mfrow = c(2,3))

for (day in 1:11) {
  Maturity = years_until_maturity
  Yield = numeric(10)
  for (bond in 1:10) {
    bond_yield <- calculate_ytm(price_data[day,bond], 100, coupons[bond],
                               years_until_maturity[bond]) * 100
    Yield[bond] <- bond_yield
  }

  plot(Maturity, Yield,
       type = "o",          # Lines and points
       col = "blue",
       lwd = 2,             # Line width
       pch = 19,            # Solid circle points
       main = "Yield Curve",
       xlab = "Maturity (Years)",
       ylab = "Yield (%)",
       ylim = c(min(Yield) * 0.9, max(Yield) * 1.1))
  mtext(bond_dates[[day,1]])
  grid()

}
```


```{r}
#| label: graph-example1
all_yields_matrix <- matrix(nrow = 10, ncol = 11)

for (day in 1:11) {
  for (bond in 1:10) {
    all_yields_matrix[bond, day] <- calculate_ytm(price_data[day, bond], 
                                                  100, coupons[bond], 
                                                  years_until_maturity[bond]) * 100
  }
}


plot(years_until_maturity, all_yields_matrix[,1], 
     type = "n",  
     main = "YTM Curves (All 11 Days)",
     xlab = "Maturity (Years)",
     ylab = "Yield (%)",
     ylim = c(min(all_yields_matrix), max(all_yields_matrix))) # Scale to fit all data
grid()

colors <- rainbow(11)
for (day in 1:11) {
  lines(years_until_maturity, all_yields_matrix[, day], 
        type = "o",       
        col = colors[day], 
        pch = 19, 
        cex = 0.8)    
}

legend("bottomright", legend = paste(bond_dates[[1]]), 
       col = colors, lty = 1, pch = 19, cex=0.4)

```

I interpolated between points with straight lines as I felt this made the least assumptions on the data between any two points. I don't really expect the real YTM to lie exactly in the middle of any two bonds; it would be difficult to conject whether it would be above or below this point.

## b) Spot Rates

### Pseudo-code
The algorithm for deriving spot rates is reasonably simple, for any day of data our bond maturing in six months is functionally a zero coupon bond, with notional equal to the sum of face value and coupon. We now have our observed price, a notional and time until stock matures, using the compounding formula, $r_1 = \left(\frac{F + C}{P} \right) -1$, we can find our spot rate for our first bond. For our n'th bond we proceed inductively, assuming we have $r_j$ for $j<n$. Rearranging our price formula, $P = \sum_{t=1}^{n-1} \frac{C}{(1+r_t)^t} + \frac{F + C}{(1+r_n)^n}$, we solve that $r_n = \left(\frac{F+C}{P - \sum_{t=1}^{n-1} \frac{C}{(1+r_t)^t}} - 1\right)^{1/n}$

```{r}
spot_rate.day <- function(x){
  # Take x day's prices
  current_prices <- as.numeric(price_data[x, ])

  # Empty vector to store our spot rates
  spot_rates <- numeric(length(current_prices))
  
  # Loop through the bonds
  for (i in 1:length(current_prices)) {
    
    # Get data for the current bond "i"
    P <- current_prices[i]
    T <- years_until_maturity[i]
    coupon_payment <- (coupons[i] * face_value) / 2
    
    if (i == 1) {
      # Base Case
      notional <- face_value + coupon_payment
      
      # Add spot to our rates
      spot_rates[i] <- 2 * ((notional / P)^(1 / (2 * T)) - 1)
    } 
    
    else {
      # Recursive Case
      pv_prior_coupons <- 0
      
      # Loop over the bonds who's spot rates we already have
      for (j in 1:(i-1)) {
        t_cashflow <- years_until_maturity[j] 
        r_spot <- spot_rates[j]               
        
        # Discount coupon using the spot rate, and add to running turtle
        pv_prior_coupons <- pv_prior_coupons + ( coupon_payment / (1 + r_spot/2)^(2 * t_cashflow) )
      }
      
      # Calculate dirty price, and use the zero coupon formula.
      dirty_price <- P - pv_prior_coupons
      
      # The notional is Face Value + Final Coupon
      notional <- face_value + coupon_payment
        
      # Solve for the spot rate at time T
      spot_rates[i] <- 2 * ((notional / dirty_price)^(1 / (2 * T)) - 1)
    }
  }
  return(spot_rates)
}
```

### Plot

```{r, include = FALSE}
par(mfrow = c(2,3))

for (i in 1:11) {
  curve_data <- data.frame(
    Maturity <- years_until_maturity,
    Yield <- spot_rate.day(i) * 100
  )

  plot(curve_data$Maturity, curve_data$Yield,
       type = "o",
       col = "blue",
       pch = 19,
       main = "Spot Rate Curve",
       xlab = "Time horizon (Years)",
       ylab = "Spot Rates (%)",
       ylim = c(min(curve_data$Yield) * 0.9,
                max(curve_data$Yield) * 1.1))
  mtext(bond_dates[[i,1]])
  grid()
}
```

```{r}
all_yields <- sapply(1:10, function(x) spot_rate.day(x) * 100)
y_min <- min(all_yields)
y_max <- max(all_yields)

plot(years_until_maturity, all_yields[,1], 
     type = "n",  
     ylim = c(y_min, y_max), 
     main = "Spot Rates for the Next 5 Years (11 Days)",
     xlab = "Time Horizon (Years)", 
     ylab = "Spot Rate (%)")

grid()


colors <- rainbow(10) 

for (i in 1:10) {
  lines(years_until_maturity, spot_rate.day(i) * 100, 
        col = colors[i],
        type = "o", 
        pch = 19, cex = 0.5) 
}


legend("bottomright", legend = paste(bond_dates[[1]]), col = colors, lty = 1, pch = 19, cex=0.4)

```

## c) Forward Rates

### Pseudo Code

To calculate the forward rate we simply have to use the formula $F_{t,t+n} = \left(\frac{(1+S_{t+n})^{t+n}}{(1+s_t)^t}\right)^{\frac{1}{n}}-1$. $S_t$ is the spot rate for time t, in this problem t = 1. t + n therefore is the difference between the maturity dates of a bond maturing in one year and a bond maturing in n years. Looping over our spot rates for 2 to 5 (although I did 1.5 to 5 in 0.5 increments since we have the data) we get our desired forward rates.

### Plot

```{r, include = FALSE}
par(mfrow = c(2,3))
x_axis <- 0.5*c(3:10)
Forward <-  numeric(8)

for (day in 1:11) {
  spot_rate <- spot_rate.day(day)

  for (rate in 3:10) {
    n <- (years_until_maturity[rate]-years_until_maturity[2])
    Forward[rate-2] <- (((1+spot_rate[rate])^years_until_maturity[rate]) / ((1 + spot_rate[2])^years_until_maturity[2]))^(1/n) - 1
  }
  Forward <- Forward * 100
  plot(x_axis, Forward,
       type = "o",          # Lines and points
       col = "blue",
       lwd = 2,             # Line width
       pch = 19,            # Solid circle points
       main = "Forward Rates",
       xlab = "t (years)",
       ylab = "Yield (%)",
       ylim = c(min(Forward) * 0.9, max(Forward) * 1.1))
  mtext(bond_dates[[day,1]])
  grid()
}
```

```{r}
plot(x_axis, Forward,
     type = "n",  
     main = "Forward Rate Curves (All 11 Days)",
     xlab = "Time t (Years)",
     ylab = "Yield (%)",
     ylim = c(min(Forward) * 0.8, max(Forward) * 1.1))
grid()

colors <- rainbow(11)
for (day in 1:11) {
  spot_rate <- spot_rate.day(day)
  
  for (rate in 3:10) {
    n <- (years_until_maturity[rate]-years_until_maturity[2])
    Forward[rate-2] <- (((1+spot_rate[rate])^years_until_maturity[rate]) / ((1+spot_rate[2])^years_until_maturity[2]))^(1/n) - 1
  }
  Forward <- Forward*100
  lines(x_axis, Forward, 
        col = colors[day],
        type = "o", 
        pch = 19, cex = 0.5)
}

legend("bottomright", legend = paste(bond_dates[[1]]), 
       col = colors, lty = 1, pch = 19, cex=0.4)


```

# Problem 5

### YTM Time Series Covariance Matrix

```{r}
YTM_matrix <- matrix(nrow = 11, ncol = 10)
for (bond in 1:10) {
  for (day in 1:11) {
    bond_yield <- calculate_ytm(price_data[day,bond], 100, coupons[bond],years_until_maturity[bond]) * 100
    YTM_matrix[day, bond] <- bond_yield
  }
}
log_diff_YTM <- diff(log(YTM_matrix))
```

Here is the covariance matrix for YTM, it has been multiplied by 10 to the power of 5 for legibility

```{r}
knitr::kable(cov(log_diff_YTM)*10^5,digits = 3)
```

### Forward Rate TS

```{r}
FR_matrix <- matrix(nrow = 11, ncol = 8)
for (day in 1:11) {
  spot_rate <- spot_rate.day(day)
  
  for (rate in 3:10) {
    n <- (years_until_maturity[rate]-years_until_maturity[2])
    FR_matrix[day, rate-2] <- (((1+spot_rate[rate])^years_until_maturity[rate]) / ((1 + spot_rate[2])^years_until_maturity[2]))^(1/n) - 1 
  }
}
log_diff_FR <- diff(log(FR_matrix* 100))
```

Here is the covariance matrix for time series, it has been multiplied by 10 to the power of 5 for legibility

```{r}
knitr::kable(cov(log_diff_FR)*10^5,digits = 3)
```

\newpage

# Problem 6

### YTM eigenvectors and eigenvalues

```{r}
vals <- eigen(cov(log_diff_YTM))$values
vectors <- eigen(cov(log_diff_YTM))$vectors
colnames(vectors) <- signif(vals, 2)
rownames(vectors) <- colnames(log_diff_YTM)

kable(round(vectors, 4))
```

### FR eigenvectors and eigenvalues
The columns are the eigenvectors in increasing eigenvalues. The respective eigenvalue is listed across the top.
```{r}
vals <- eigen(cov(log_diff_FR))$values
vectors <- eigen(cov(log_diff_FR))$vectors
colnames(vectors) <- signif(vals, 2)
rownames(vectors) <- colnames(log_diff_FR)

kable(round(vectors, 4))
```

### Comments

This vector in bond space points in the direction where the data is most spread out, that is the weighted combination of stocks that explains the most risk. The size of it's eigenvalue, compared to the others tells us how much of the variation is explained by this portfolio.

# Assumptions, Tables, Graphs

It was assumed prices were quoted on the first of January, and that every year has 365 days. For discounting it was assumed that the final coupon payment made at the same date of maturity (together with face value). 

# References

Milstein, E., & Wessel, D. (n.d.). What did the Fed do in response to the COVID-19 crisis? Brookings. Retrieved January 14, 2026, from https://www.brookings.edu/articles/fed-response-to-covid19/

Markets Insider. (n.d.-a). Bond finder [United States Treasury short-term bond search results]. Business Insider. Retrieved January 19, 2026, from https://markets.businessinsider.com/bonds/finder?borrower=71&maturity=shortterm&yield=&bondtype=2%2c3%2c4%2c16&coupon=&currency=184&rating=&country=19

Markets Insider. (n.d.-b). Bond finder [United States Treasury mid-term bond search results]. Business Insider. Retrieved January 19, 2026, from https://markets.businessinsider.com/bonds/finder?borrower=71&maturity=midterm&yield=&bondtype=2%2c3%2c4%2c16&coupon=&currency=184&rating=&country=199
